---
title: Direct Transmission Modes - Practice
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null   
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsaide_references.bib
---


```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
This app allows you to explore a model with 2 different modes of direct transmission. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.


## The Model {#shinytab2}


### Model Overview

This is a simple SIR model with births and deaths and 2 types of direct transmission, density-dependent and frequency-dependent (based on the terminology in [@begon02] - see _Further Information_ tab for references).

This model has the following compartments:  

* **S** - uninfected and susceptible individuals 
* **I** - individuals who are infected and infectious.
* **R** - recovered/removed individuals.

The processes being modeled are:

* Susceptible individuals can become infected by two types of direct transmission from infected hosts, further described below.
* Infected hosts recover after some time (specified by the rate _g_). 
* Recovered individuals are initially immune to reinfection. They can lose their immunity at rate _w_ and return to the susceptible compartment. 
* New susceptibles enter (are born) at a rate _n_. From each compartment, hosts "leave" after some time (i.e. they die) at rate _m_. The inverse of this rate is the average lifespan of a host. 



The force of infection, denoted by _f_ here, depends on the transmission model. For density-dependent transmission, it is given by _f = b~d~ I / A_, where _A_ is the area of the scenario under consideration (e.g. a city). For frequency-dependent transmission, we have _f = b~f~ I / N_, where _N_ is the size of the population, _N=S+I+R_. 



### Model Implementation
The flow diagram and the set of ordinary differential equations (ODE) which are used to implement this model are as follows:


```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```



$$\dot S =n - f S  + wR - mS$$ 
$$\dot I = f S - g I - mI$$ 
$$\dot R = g I - wR - mR$$

with _f = b~d~ I / A_ for density-dependent transmission and _f = b~f~ I / N_ for frequency-dependent transmission.


### Model Comments
Note that quite often, we assume that the area is constant and do not include it in the model or even talk about it. Often, the population size is also assumed to be constant. In that case, the 2 types of transmission models behave the same. But as soon as area or population size change, the model results differ. You can explore the effect of this relationship using this simulation.

## What to do {#shinytab3}


**The tasks below are described in a way that assumes everything is in units of MONTHS (rate parameters, therefore, have units of inverse months). If any quantity is not given in those units, you need to convert it first (e.g. if it says a year, you need to convert it to 12 months).**

```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 5 #number of tasks
nrecord = 6 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Run the simulation with 1000 susceptible individuals and 1 initially infected host. Set the simulation duration, _tmax_ to 10 years. Consider density-dependent transmission, with a transmission rate of _b~d~_ = 0.005, and an area of size 2. Assume that the duration of the infectious period is 15 days long (and that there are 30 days in a month). Turn off births, deaths and waning immunity for now. If you did it correctly, your outbreak should end with around 625 susceptible individuals still remaining."

# Record for task 1
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Total number of recovered at end of outbreak")
reclist$rectype = "Numeric"
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)

allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Now switch to frequency-dependent transmission, set _b~f~_ = 2.5. Leave all other settings as before.
B) Think about your expectations, run the simulation and compare your expectations with the results.Anything surprising happening? Do you understand why you see what you see?"

# Record for task 2
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 3 
tid = 3
alltasks[tid,"TaskText"] = "A) Let's assume we are now in a location with twice the number of people as before (_S~0~_ = 1999) , living in the same area. Implement that in the simulation by changing population size, keep all other settings as above.
B) What do you expect to see for the frequency and density dependent scenarios? Does anything change? Do you understand why?"

# Record for task 3
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Total number of recovered at end of outbreak (scenario = 1)",
                    "Total number of recovered at end of outbreak (scenario = 2)")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec

### Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) If you double the population size as you just did, how do you need to adjust the area to obtain the same sized outbreak with regard to the fraction of people getting infected/remaining susceptible for density-dependent transmission? Try with the simulation and see if your expectation is correct."

# Record for task 4
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("The area used to get the same fraction")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) Keep exploring by trying different parameters and transmission settings and see how they influence results. You can also go beyond a single outbreak and turn on births/deaths (which can impact population size) or waning immunity.
B) As you continue your exploration, think about real infectious diseases that might be approximated by either one of the transmission types, and what approximate choices for the model parameters would describe those IDs."

# Record for task 5
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```



## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* For more on this topic, see e.g. [@begon02]. 

### References

