---
title: Evolutionary Dynamics - Practice
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsaide_references.bib
---


```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
For this module, we will explore a stochastic SIR-type model with 2 different pathogen strains, wild-type and a drug resistant mutant in the presence of drug treatment. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.

### Learning Objectives


* Understand the central mechanisms and drivers of infectious disease evolution
* Be able to assess how different control strategies can affect infectious disease evolution


## The Model {#shinytab2}

### Model Overview

This model tracks susceptibles, wild-type infected untreated, wild-type infected treated, drug-resistant infected and recovered hosts. 
The following compartments are included:  

* __S__ - uninfected and susceptible individuals 
* __I~u~__ - individuals who are infected with the wild-type/drug sensitive strain and not on treatment.
* __I~t~__ - individuals who are infected with the wild-type/drug sensitive strain and on treatment.
* __I~r~__ - individuals who are infected with the drug-resistant strain. Treatment has no impact on those hosts.
* __R__ - recovered/removed individuals. Those individuals have recovered and are immune. 


The included processes/mechanisms are the following:

* Susceptible individuals, _S_ can become infected by untreated or treated hosts infected with the wild-type strain at rates _b~u~_ and _b~t~_. 
* A fraction _f_ of those hosts infected with the wild-type strain are assumed to receive treatment (_I~t~_), while the remainder does not (_I~u~_). 
* Small fractions, that is low values for _c~u~_ and _c~t~_  of wild-type infected untreated or treated hosts can generate resistant mutants and transmit those, leading to resistant infections.
* Individuals infected with the resistant strain ,_I~r~_, might receive treatment, but we assume treatment has no effect in those individuals so don't track them explicitly.
* Resistant infected individuals can infect susceptible individuals (S) with the resistant strain at rate _b~r~_. 
* All infected hosts recover at some rate _g~i~_, with _i=u,t,r_. 
* Recovered individuals are immune to reinfection. 



### Model Implementation
The flow diagram for the model implemented in this app is:

```{r modeldiagram,  fig.cap='',  echo=FALSE, out.width = "70%"}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```

The deterministic/ODE equations for this model are:

$$\dot S = - S (b_u I_u + b_t I_t + b_r I_r)$$ 
$$\dot I_u = S (1-f) b_u (1-c_u) I_u + S (1-f)b_t (1-c_t) I_t  - g_u I_u$$
$$\dot I_t = S f b_u (1-c_u) I_u + S f b_t (1-c_t) I_t - g_t I_t$$
$$\dot I_r = S (b_u c_u I_u + b_t c_t I_t + b_r I_r) - g_r I_r$$
$$\dot R = g_u I_u + g_t I_t + g_r I_r$$

However, for this app we do __not__ implement a deterministic/ODE model. Instead, we implement its stochastic equivalent.
We can specify the model by writing down every possible transition/event/reaction that can occur and their propensities (the propensity multiplied with the
time step gives the probability that a given event/transition occurs). For our model these are the following:

Event type        |  Transitions               |   Propensity      |
----------        |  -----------               |   ----------      |
S turn into I~u~  |  S => S-1,  I~u~ => I~u~+1 |   (1-f) * (b~u~ * (1-c~u~) * I~u~ + b~t~ * (1-c~t~) * I~t~) * S  |
S turn into I~t~  |  S => S-1,  I~t~ => I~t~+1 |   f * (b~u~ * (1-c~u~) * I~u~ + b~t~ * (1-c~t~) * I~t~) * S  |
S turn into I~r~  |  S => S-1,  I~r~ => I~r~+1 |   (b~u~ * c~u~ * I~u~ + b~t~ * c~t~ * I~t~ + b~r~ * I~r~) * S  |
Recovery of I~u~  |  I~u~ => I~u~-1, R => R+1  |   g~u~ * I~u~       |
Recovery of I~t~  |  I~t~ => I~t~-1, R => R+1  |   g~t~ * I~t~       |
Recovery of I~r~  |  I~r~ => I~r~-1, R => R+1  |   g~r~ * I~r~       |




## What to do {#shinytab3}


**The tasks below are described in a way that assumes everything is in units of DAYS (rate parameters, therefore, have units of inverse days). If any quantity is not given in those units, you need to convert it first (e.g. if it says a week, you need to convert it to 7 days).**


```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 6 #number of tasks
nrecord = 18 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Set the model parameters to the following: susceptible _S~0~_ = 500, and initially untreated infected host _I~u0~_ = 1. No other infected hosts _I~t0~_ = 0 and _I~r0~_ = 0. Set simulation duration, _tmax_, to approximately half a year. Assume that untreated individuals transmit at _b~u~_ = 0.001, treated at _b~t~_ = 0.0005, and resistant at _b~r~_ = 0.0008. Assume that the duration of the infectious period is 5 days for untreated, 4 days for treated and 5 days for resistant (for those individuals, treatment has no effect). Set the rates _g~i~_ accordingly. Set the number of simulations to 20. Assume nobody receives treatment and no resistance is generated (_f = c~u~ = c~t~ = 0_). With parameters set to correspond to the scenario just described, run the simulation. You should see some simulations with outbreaks and some without. For those _with outbreaks_, you should have around 10-100 susceptible left at the end."

# Record for task 1
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) With everything as before, set the initial number of untreated infected hosts to 10. Run the simulations. You should pretty much always get outbreaks. Why do you almost always have outbreaks for this setting and fewer outbreaks/more extinctions for task 1?"

# Record for task 2
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("The average number of susceptibles left")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


# Task 3
tid = 3
alltasks[tid,"TaskText"] = "A) With the same settings as _Task 2_ turn on treatment ( _f_ > 0 ). Set number of simulations to 50 (the simulations might take a few seconds to run). Run the simulation with fraction receiving treatment, _f_, at 0, 0.25, 0.5, 0.75 and 1. Draw conclusions about the usefulness of treatment."

# Record for task 3
nrec = 5 # number of items to record
reclist = list()
reclist$rectext = c("The average number of susceptibles at the end (_f_ = 0)",
                    "The average number of susceptibles at the end (_f_ = 0.25)",
                    "The average number of susceptibles at the end (_f_ = 0.5)",
                    "The average number of susceptibles at the end (_f_ = 0.75)",
                    "The average number of susceptibles at the end (_f_ = 1)")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  


# Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Now allow resistance to be generated during treatment (_c~t~_ > 0). Set _c~t~ = 0.2_ for the fraction of resistant generation from treatment. Run the simulation for the treatment levels specified in the previous task. Draw conclusions about the usefulness of treatment. In your head or on a piece of paper, sketch out the relationship between treatment level and the number of susceptibles left at the end in the absence and presence of resistance generation (_c~t~_ = 0 and _c~t~_ > 0). What do you conclude from that?"

# Record for task 4
nrec = 5 # number of items to record
reclist = list()
reclist$rectext = c("The average number of susceptibles at the end (_f_ = 0)",
                    "The average number of susceptibles at the end (_f_ = 0.25)",
                    "The average number of susceptibles at the end (_f_ = 0.5)",
                    "The average number of susceptibles at the end (_f_ = 0.75)",
                    "The average number of susceptibles at the end (_f_ = 1)")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  


# Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) Set the rate of transmission for resistant hosts to _b~r~ = 0.001_. Keep untreated hosts and treated host values as before. Set resistance generation during treatment to _c~t~ = 0.3_.  Keep everything else as previously. Contemplate what these changes mean biologically, and what you should expect from the simulations. Run the model for each of the 5 treatment levels specified above and record the average number of susceptibles left at the end. Again, sketch out the relationship between treatment level and the number of susceptibles left at the end and compare to what you found in tasks 3 and 4. What do you conclude from that?"

# Record for task 5
nrec = 5 # number of items to record
reclist = list()
reclist$rectext = c("The average number of susceptibles at the end (_f_ = 0)",
                    "The average number of susceptibles at the end (_f_ = 0.25)",
                    "The average number of susceptibles at the end (_f_ = 0.5)",
                    "The average number of susceptibles at the end (_f_ = 0.75)",
                    "The average number of susceptibles at the end (_f_ = 1)")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

# Task 6
tid = 6
alltasks[tid,"TaskText"] = "A) Keep exploring. For instance, try the following: Turn on resistance generation for both treated, _c~t~_, and untreated, _c~u~_. Explore how population size, _S~0~_, the fraction of resistance generation ( _c~t~_ or _c~u~_ ) and fitness of the different strains ( _b~t~_ or _b~u~_ or _b~r~_ ) affect outcomes. If you want to explore the relation between treatment level and outbreak size (tasks 3-5) in more detail, you could access the underlying simulation directly and write a bit of R code to loop over treatment values following the _Level 2_ example described in the DSAIDE tutorial."

# Record for task 6
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```

      
## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* The model is a simpler version of a previously published model in [@handel09a].


### References
