---
title: Host Heterogeneity - Practice
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: references.bib
---



```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
For this module, we will explore a simple SIR model with 2 types of hosts. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.

## The Model {#shinytab2}

### Model Overview

This model tracks susceptibles, infected and recovered of 2 different types. Think of those types as e.g. males/females, children/adults, etc. 

The following compartments are included, twice for each type (i=1,2):  

* **S~i~** - uninfected and susceptible individuals 
* **I~i~** - individuals who are infected and infectious.
* **R~i~** - recovered/removed individuals. Those individuals have recovered and are immune. 


The included processes/mechanisms are the following:

* Susceptible individuals of type 1 or 2 can become infected by infected hosts of type 1 or 2. The rates at which infections to type 1/2 from type 1/2 occur are given by the parameters _b~11~_, _b~12~_, _b~21~_ and _b~22~_. 
* Infected hosts of type 1/2 recover after some time at rates _g~1~_/_g~2~_. 
* Recovered individuals are initially immune to reinfection. They can lose their immunity at rates _w~1~_/_w~2~_ and return to the susceptible state. 
 


### Model Implementation
The flow diagram and equations describe the model implemented in this app:

```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```


$$\dot S_1 = - S_1 (b_{11} I_1 + b_{12} I_2) + w_1 R_1 $$ 
$$\dot I_1 =  S_1 (b_{11} I_1 + b_{12} I_2) - g_1 I_1 $$ 
$$\dot R_1 = g_1 I_1 - w_1 R_1 $$
$$\dot S_2 = - S_2 (b_{22} I_2 + b_{21} I_1) + w_2 R_2 $$ 
$$\dot I_2 =  S_2 (b_{22} I_2 + b_{21} I_1) - g_2 I_2  $$ 
$$\dot R_2 = g_2 I_2 - w_2 R_2 $$




## What to do {#shinytab3}


**The tasks below are described in a way that assumes everything is in units of MONTHS (rate parameters, therefore, have units of inverse months). If any quantity is not given in those units, you need to convert it first (e.g. if it says a year, you need to convert it to 12 months).**

```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 15 #number of tasks
nrecord = 25 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Set the model to the following 1000 susceptible hosts and 1 infected host of type 1. 200 susceptible hosts and 1 infected host of type 2. Simulation duration approximately 5 years. Assume that transmission from host 1 to host 1 is _b~11~ = 0.002_, from host 2 to host 2 is _b~22~ = 0.01_. No transmission from one host type to the other _b~12~ = 0_ and _b~21~ = 0_. Assume that the duration of the infectious period is 1 month long for both types ( _g~1~_ and _g~2~_ ). No waning immunity. Run the simulation and ensure you get outbreaks in both populations with 20% susceptibles left at the end."

# Record for task 1
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Set _b~11~ = 0.001_. Rest as before. Run the simulation. You should get the same outbreak as before among type 2 hosts, no real outbreak among type 1 hosts. If you have worked through the reproductive number app already, contemplate what R~0~ was for both types/populations in task 1, what it is now in task 2, and how that influences what you see in the simulation."

# Record for task 2
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("R~0~ for type 1 hosts",
                    "R~0~ for type 2 hosts")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Round to the tenths",
                    "Round to the tenths")
reclist$recfuzzy = c(0.1,
                     0.1)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


# Task 3
tid = 3
alltasks[tid,"TaskText"] = "A) Now set the transmission rate from host 2 to host 1 _b~21~ = 0.001_. Everything as before. Run the simulation. You should see an outbreak in both populations. This is an example of a core group driving the dynamics. Contemplate what the core group does."

# Record for task 3
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  


# Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Set initial number of infected type 1 hosts to 0. Leave all other settings as before. Contemplate what you expect to see, run the simulation, see if your expectations are confirmed."

# Record for task 4
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task


# Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) Now set the initial number of infected type 1 hosts back to 1, and infected type 2 hosts to 0. Contemplate what you expect to see, run the simulation, see if your expectations are confirmed."

# Record for task 5
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

# Task 6
tid = 6
alltasks[tid,"TaskText"] = "A) Set _I1~0~ = 1_ and _I2~0~ = 1_. Set the transmission rates to host 1 from host 1 and to host 2 from host 2 to 0. Set the transmission rate to host 1 from host 2  to _b~12~_ = 0.01, to host 2 from host 1  to _b~21~_ = 0.002. Those are the same values used in task 1 for transmission among the same hosts. Contemplate what you expect to see, run the simulation, see if your expectations are confirmed. Transmission only between hosts of different type could for instance represent a sexually transmitted disease in a heterosexual population, with the 2 types of hosts being females and males."

# Record for task 6
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 recovered at the end",
                    "Number of type 2 recovered at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task


# Task 7
tid = 7
alltasks[tid,"TaskText"] = "A) Now set the transmission rate _b~12~ = 0.01_, and _b~21~ = 0.002_. Everything else as before. Again, think about your expectations, run simulation and check."

# Record for task 7
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 recovered at the end",
                    "Number of type 2 recovered at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task

# Task 8
tid = 8
alltasks[tid,"TaskText"] = "A) Repeat the previous task, with only an infected of type 1 initially present. Contemplate what you expect to see, run the simulation, see if your expectations are confirmed."

# Record for task 8
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 9
tid = 9
alltasks[tid,"TaskText"] = "A) Repeat the previous task, with only an infected of type 2 initially present. Contemplate what you expect to see, run the simulation, see if your expectations are confirmed."

# Record for task 9
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

# Task 10
tid = 10
alltasks[tid,"TaskText"] = "A) Set everything back as in task 1, but with waning immunity with an average duration of immunity of 5 months for each population (i.e. rates _w~1~_ and _w~2~_ need to be the inverse of 5 months.). Run simulation, confirm that both populations reach a steady endemic state."

# Record for task 10
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 11
tid = 11
alltasks[tid,"TaskText"] = "A) Set transmission rate _b~12~ = 0.005_, everything else as before. Run the simulation and see what changes."

# Record for task 11
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 12
tid = 12
alltasks[tid,"TaskText"] = "A) Set transmission rate _b~21~ = 0.005_, and _b~12~ = 0_, everything else as before. Run the simulation and see what changes."

# Record for task 12
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 13
tid = 13
alltasks[tid,"TaskText"] = "A) With the settings you just had, turn off waning immunity for type 1, run the simulation and see what you get."

# Record for task 13
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

# Task 14
tid = 14
alltasks[tid,"TaskText"] = "A) Turn waning immunity for type 1 back on and turn off for type 2, run the simulation and see what you get."

# Record for task 14
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Number of type 1 susceptible remaining at the end",
                    "Number of type 2 susceptible remaining at the end")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task   

# Task 15
tid = 15
alltasks[tid,"TaskText"] = "A) Keep exploring. Think of a real ID where accounting for 2 types of hosts is important, see if you can use the simulation to approximate and explore that ID."

# Record for task 15
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 
```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```



## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* For some more information dealing with host heterogeneity, see e.g. [@beldomenico10, @yorke78, @lloyd-smith05].

### References
