---
title: ID Control 1 - Practice
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: references.bib
---


```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```




## Overview {#shinytab1}
This app links the reproductive number and ID control. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab. Before going through this app, you should go through the 'Reproductive Number' apps first.

### Learning Objectives
* Know how to the reproductive number informs control. 
* Know how different control measures impact the reproductive number.


## The Model {#shinytab2}

### Model Overview
For this app, we'll use the same basic compartmental SIR model as for the 'Reproductive Number 2' app. We allow for 3 different stages/compartments: 

* **S** - uninfected and susceptible individuals 
* **I** - infected and infectious individuals (note that these terms are often used interchangeably, but technically we are talking about someone who is infected **and** is infectious, i.e. can infect others)
* **R** - recovered/removed individuals. Those are individuals that do not further participate, either because they are now immune or because they died.


In addition to specifying the *compartments* of a model, we need to specify the dynamics determining the changes for each compartment. Broadly speaking, some processes increase the number of individuals in a given compartment/stage and other processes that lead to a reduction. Those processes are sometimes called inflows and outflows. 

For our system, we specify the following processes/flows: 

* Susceptible individuals (S) can become infected by infectious individuals (I) at some rate, _b_. This leads to the susceptible individual leaving the S compartment and entering the I compartment.   
* Infected individuals recover and enter the recovered (R) compartment at some rate, _g_.
* Recovered individuals can lose immunity at rate _w_ and return to the susceptible compartment. 
* Natural births (into the S compartment at rate _n_) and deaths (from all compartments at rate _m_) are possible.
* The model allows a fraction of individuals to be vaccinated before the outbreak. Those individuals are moved into the R compartment before the start of the outbreak.



### Model Implementation
The flow diagram and the set of equations which are used to implement this model are as follows:


```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```



$$\dot S =n - b SI - mS + wR$$ 
$$\dot I = b S I - g I - mI$$
$$\dot R = g I - mR - wR$$
$$S_{v} = (1-ef)S(0)$$
$$R(0) = efS(0)$$




Here, _S(0)_ is the initial population of susceptibles, and _S~v~_ is the susceptible population left after vaccinating a fraction _f_ (at vaccine efficacy _e_). Vaccinated and protected individuals are moved to the _R_ compartment before the start of the outbreak, _R(0)_. Note the unfortunate fact that the recovered compartment uses the same letter as the reproductive number, and the starting value for the _R_ compartment, _R(0)_ looks similar to the basic reproductive number. This is common notation and I therefore use it here too. Just be careful to make sure you know which quantity is discussed.



## What to do {#shinytab3}


**The tasks below are described in a way that assumes everything is in units of MONTHS (rate parameters, therefore, have units of inverse months). If any quantity is not given in those units, you need to convert it first (e.g. if it says a year, you need to convert it to 12 months).**

```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 4 #number of tasks
nrecord = 6 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Knowing the reproductive number, _R_, is important for control strategies, e.g. for vaccine campaigns. You learned in the reproductive number apps that for _R=1_ an outbreak switches from growth to decline (often called the threshold value). Let's say you have an ID that enters a new population where everyone is susceptible. That ID has R~0~=4. Would you expect to see an outbreak? Why? Now let's assume that we protected half the population through a (100% effective) vaccine. What is the new value for R, i.e. how many people are being infected on average by an infected person after we vaccinated? Is that new value of R low enough to prevent the outbreak? How low does R need to be to prevent an outbreak? What fraction of the population would you need to protect/vaccinate to achieve such an R?"

# Record for task 1
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("The new value for R",
                    "The proportion vaccinated needed to prevent an outbreak")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report integer",
                    "Report as decimal")
reclist$recfuzzy = c(0,
                     0.01)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Let's test the vaccination idea with the computer simulation. Set the simulation with 1000 susceptibles and 1 infected, simulation time 24 months, _g_=5, no births, deaths or waning immunity. Choose the value for _b_ such that _R~0~=4_. Run the simulation for 0% vaccination coverage to confirm things happen as you expect. Check if the final size of the susceptible/infected population is what it should be and returns (approximately) an R~0~=4 when you use the final size equation to compute it. 

B) Now set 50% vaccination coverage at 100% vaccine efficacy (_f=0.5_ and _e=1_). What is the value of the effective _R_ after vaccination? Run a simulation and check to make sure that is what you get. Finally, run the simulation at the vaccination level you determined above to be enough to prevent an outbreak. Make sure your expectations and simulation results agree."

# Record for task 2
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Effective _R_ after vaccination (Part B, _f=0.5_)")
reclist$rectype = "Numeric"
reclist$recnote = c("Report two decimals")
reclist$recfuzzy = c(0.01)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 3
tid = 3
alltasks[tid,"TaskText"] = "A) Most vaccines are not perfect. What fraction of the population would you need to vaccinate to prevent an outbreak if the vaccine efficacy/effectiveness was 75%? What if it was 60%? Confirm with the simulations."

# Record for task 3
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("The proportion vaccinated needed to prevent an outbreak (_e=.75_)",
                    "The proportion vaccinated needed to prevent an outbreak (_e=.60_)")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report two decimals",
                    "Report two decimals")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Another useful intervention is quarantine, these days more commonly referred to as social distancing. By either forcibly or voluntarily reducing contacts between susceptibles and infected, this can reduce the transmission rate and thus the reproductive number. Consider the scenario as in task 2, but now we have a _b=0.015_. If we were able to reduce contacts and thus transmission by half, would that be enough to prevent an outbreak? By what fraction do we need to reduce the transmission to prevent an outbreak? To answer this, first compute the corresponding _R~0~_. Then confirm with the model. 

Note: In this model, we can't distinguish between interventions that target susceptibles or infected, both could reduce the transmission rate. Targeting susceptibles before they become infected is of course preferable, but there are usually many more of those, so targeting infected is often easier. Interventions that reduce contact and transmission/infection risk for both groups is of course best."

# Record for task 4
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Value of _b_ that prevents an outbreak")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report decimal to thousandths")
reclist$recfuzzy = c(0.001)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```



## Further Information {#shinytab4}
* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* For the equations mentioned in the tasks, see e.g. [@heffernan05a; @keeling08; @vynnycky10].
* For some information on the reproductive number, see e.g. [@fine11; @heffernan05a; @roberts07; @wallinga07].

### References



