---
title: Control applied to different populations 
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: references.bib
---

```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}  
This app lets you explore interventions applied to different population subgroups with potentially different infectiousness and mortality patterns. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab. Before going through this app, it might be useful to go through the 'Reproductive Number' and `Multi Outbreak ID Control` apps first. This app closely follows a model and analysis described in [@Handel2020], see the "Further Information" tab for this and other related references.


### Learning Objectives  
* Understand the concept of interventions applied to specific population groups.  
* Learn the impact of different interventions on outcomes such as  mortality.
* Explore how optimal control measures depend on characteristics and composition of a population.  


## The Model {#shinytab2}

### Model Overview
For this app, we use a version of the basic compartmental SIR-type model. We track the following stages:

* **$S_i$** - uninfected and susceptible individuals 
* **$I_i$** - infected and infectious individuals
* **$R_i$** - recovered individuals
* **$D_i$** - dead individuals

Individuals are divided into 3 populations, namely children, adults and elderly ($i=c,a,e$). Thus, our model with 3 age groups and 4 compartments per age group leads to a 12-compartment model.

For this app, we specify the following processes/flows: 

* $b_{ij}$ is the rate of transmission to susceptible individuals ($S_i$) from infected individuals ($I_j$). This leads to the susceptible individual leaving the S compartment and entering the I compartment. Transmission among and between groups can occur at different rates.   
* Infected individuals stay infectious for some time, and leave the infected/infectious compartment at rate $g_i$. A fraction $m_i$ of those individuals will die and enter the dead compartment, the remaining fraction $1-m_i$ will enter the recovered compartment. 
* Recovered individuals are initially immune to re-infection, but can lose immunity at rate $w_i$.
* Control measures can applied to each group at some strength $f_i$ (between 0 and 1) which reduces the rate at which infections in that group occur. Control measures for each group start at time $f^i_{start}$ and end at time $f^i_{end}$.


### Model Implementation
The flow diagram and the set of equations which are used to implement this model are as follows:

```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```


Model equations:

$$
\begin{aligned}
\dot{S}_c &= - (1 - f_1)(b_{cc} I_c + b_{ca} I_a + b_{ce} I_e) S_c + w_c R_c\\
\dot{I}_c &=   (1 - f_1)(b_{cc} I_c + b_{ca} I_a + b_{ce} I_e) S_c - g_c I_c\\
\dot{R}_c &= (1-m_c) g_c I_c - w_c R_c\\
\dot{D}_c &= m_c g_c I_c\\
\dot{S}_a &= - (1 - f_2)(b_{ac} I_c + b_{aa} I_a + b_{ae} I_e) S_a + w_a R_a\\
\dot{I}_a &=   (1 - f_2)(b_{ac} I_c + b_{aa} I_a + b_{ae} I_e) S_a - g_a I_a\\
\dot{R}_a &= (1-m_a) g_a I_a - w_a R_a\\
\dot{D}_a &= m_a g_a I_a\\
\dot{S}_e &= - (1 - f_3)(b_{ec} I_c + b_{ea} I_a + b_{ee} I_e) S_e + w_e R_e\\
\dot{I}_e &=   (1 - f_3)(b_{ec} I_c + b_{ea} I_a + b_{ee} I_e) S_e - g_e I_e\\
\dot{R}_e &= (1-m_e) g_e I_e - w_e R_e\\
\dot{D}_e &= m_e g_e I_e
\end{aligned}
$$


## What to do {#shinytab3}

In the following, we explore a few simple hypothetical interventions for different scenarios and assess their effects.

```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 5 #number of tasks
nrecord = 13 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) We start by considering an outbreak of something like COVID-19. We assume a population of 1 million individuals, with 20% children, 25% elderly and the remainder middle-aged adults. Set the values for $S_c$, $S_a$ and $S_e$ to reflect these assumptions. We consider rates of transmission to be larger within groups than between groups. We also choose values to have an overall reproductive number of 2. This can be achieved by setting $b_{cc} = 3*10^{-7}$, $b_{ca} = 1.5*10^{-7}$, $b_{ce} = 7.5*10^{-8}$, $b_{ac} = 1.5*10^{-7}$, $b_{aa} = 3*10^{-7}$, $b_{ae} = 1.5*10^{-7}$, $b_{ec} = 7.5*10^{-8}$,  $b_{ea} = 1.5*10^{-7}$, $b_{ee} = 2.25*10^{-7}$. We make the assumption that infection lasts around 10 days in each age group. This means setting the rates $g_i=0.1$. To obtain age-related mortality similar to COVID, we assume mortality risks of 0.1%, 1% and 10% for children, adults and elderly respectively. Set the parameters $m_i$ accordingly. Run the simulation without any control applied, i.e. $f_1 = f_2 = f_3 = 0$. You should see an outbreak with susceptibles at the end of the outbreak at $S_c = 56799$, $S_a = 84498$, $S_e = 75413$ and deaths of $D_c = 143$, $D_a = 4655$, $D_e = 17459$."

# Record for task 1
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Now lets implement strong control for each age group by setting $f_1 = f_2 = f_3 = 0.9$. Assume that each control starts 90 days after outbreak start and lasts for 90 days. Keep all other values as previously. Run the simulation. You should see that the strong control initially pushes infected down, but once control is stopped, the outbreak rebounds and the overall size of the outbreak is similar to that without control, with susceptibles at the end of the outbreak at $S_c = 63247$, $S_a = 99107$, $S_e = 83557$"

# Record for task 2
nrec = 3 # number of items to record
reclist = list()
reclist$rectext = c("Number of D_c at the end",
                    "Number of D_a at the end",
                    "Number of D_e at the end")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task   


# Task 3
tid = 3
alltasks[tid,"TaskText"] = "A) Now lets implement control that targets mainly children. We also need to apply some control to adults, the largest group, to make the effective reproductive number values comparable. This can be achieved by setting $f_1=1; f_2=0.339; f_3=0$. Assume that each control starts 90 days after outbreak start and lasts for 90 days. Keep all other values as previously. Run the simulation. You should see susceptibles at the end of the outbreak at $S_c = 120683$, $S_a = 159146$, $S_e = 91818$"

# Record for task 3
nrec = 3 # number of items to record
reclist = list()
reclist$rectext = c("Number of D_c at the end",
                    "Number of D_a at the end",
                    "Number of D_e at the end")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task    


# Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Now lets implement control applied to adults only, in such a way that the effective reproductive number is the same as for the previous task. This can be achieved by setting $f_1=0; f_2=0.475; f_3=0$. Keep all other values as previously. Run the simulation. You should see susceptibles at the end of the outbreak at $S_c = 69000$, $S_a = 181882$, $S_e = 91279$"

# Record for task 4
nrec = 3 # number of items to record
reclist = list()
reclist$rectext = c("Number of D_c at the end",
                    "Number of D_a at the end",
                    "Number of D_e at the end")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  


# Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) Now lets implement control that targets mainly the elderly group, again keeping the overall effective reproductive number the same. This can be achieved by setting $f_1=0; f_2=0.318; f_3=1$. Keep all other values as previously. Run the simulation. You should see susceptibles at the end of the outbreak at $S_c = 69988$, $S_a = 157328$, $S_e = 155342$"

# Record for task 5
nrec = 3 # number of items to record
reclist = list()
reclist$rectext = c("Number of D_c at the end",
                    "Number of D_a at the end",
                    "Number of D_e at the end")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task   

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```


## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course, to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* This app closely follows a model and analysis described in [@Handel2020]. 

### References

