---
title: Complex ID control Scenarios - Practice
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsaide_references.bib
---


```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
For this module, we will explore a fairly complex model that allows the study of different types of interventions. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.

## The Model {#shinytab2}

### Model Overview

This model is fairly big and has many parts that can be turned on or off depending on parameter settings. The model allows for 3 types of transmission: direct, through an environmental stage, and through a vector stage. The (human) host is modeled in some detail, the environment and vectors are modeled with 1 and 2 compartments. 
The following compartments are included:  

* **S** - uninfected and susceptible individuals 
* **P** - individuals who are infected and do not yet show symptoms. Those individuals can potentially be infectious
* **A** - individuals who are infected and do not show symptoms. Those individuals can potentially be infectious
* **I** - individuals who are infected and show symptoms. Those individuals are likely infectious, but the model allows to adjust this, including no infectiousness.
* **R** - recovered/removed individuals. Those individuals have recovered and are immune. They can loose their immunity in this model.
* **D** - individuals who have died due to the disease.
* **E** - pathogen in the environment
* **S~V~** - susceptible vectors
* **I~V~** - infected/infectious vectors


The included processes/mechanisms are the following:

* Susceptible individuals (S) can become infected by pre-symptomatic (P), asymptomatic (A) or symptomatic (I) hosts. The rates at which infections from the different types of infected individuals (P, A and I) occur are governed by 3 parameters, _b~P~_, _b~A~_, and _b~I~_. 
* Susceptible individuals (S) can also become infected by contact with the environment or infected vectors, at rates _b~E~_ and _b~v~_. 
* Susceptible vectors (S~v~) can become infected by contact with symptomatic hosts at rate _b~h~_. 
* All infected hosts first enter the presymptomatic stage. They remain there for some time (determined by rate _g~P~_, the inverse of which is the average time spent in the presymptomatic stage). 
A fraction _f_ of presymptomatic hosts move into the asymptomatic category, and the rest become symptomatic infected hosts.
* Asymptomatic infected hosts recover after some time (specified by the rate _g~A~_). Similarly, the rate _g~I~_ determines the duration the symptomatic hosts stay in the symptomatic state. For symptomatic hosts, two outcomes are possible. Either recovery or death. The parameter _d_ determines the fraction of hosts that die due to disease.
* Recovered individuals are initially immune to reinfection. They can lose their immunity at rate _w_ and return to the susceptible compartment. 
* Symptomatic and asymptomatic hosts shed pathogen into the environment at rates p~A~ and p~I~. The pathogen in the environment decays at rate _c_.
* New susceptible hosts and vectors enter the system (are born) at rates _n~h~_ and _n~v~_. Mortality (death unrelated to disease) for hosts and vectors occurs at rates _m~h~_ and _m~v~_.

Note that we only track people that die due to the disease in our _D_ compartment. All hosts dying due to other causes just "exit the system" and we don't further keep track of them (though we could add another compartment to "collect" and track all individuals who died from non-disease-related causes.)

Also, note that we made several simplifications to keep the model from getting too complex. For instance, presymptomatic individuals do not shed into the environment, and only symptomatic hosts are assumed to be able to infect vectors. Further details relaxing these assumptions could, of course, be included, at the expense of a larger and more complex model.


### Model Implementation
The flow diagram and equations describe the model implemented in this app (Note: births and natural deaths are not drawn to keep the diagram from getting too cluttered):


```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE, out.width = "70%"}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```


$$\dot S = n_h - S (b_P P + b_A A + b_I I + b_E E + b_v I_v) + wR - m_h S $$ 
$$\dot P = S (b_P P + b_A A + b_I I + b_E E + b_v I_v) - g_P P - m_h P$$ 
$$\dot A = f g_P P - g_A A - m_h A$$ 
$$\dot I = (1-f) g_P P - g_I I - m_h I $$ 
$$\dot R = g_A A + (1-d) g_I I - wR - m_h R$$
$$\dot D = d g_I I $$
$$\dot E = p_I I + p_A A - c  E $$
$$\dot S_v = n_v  - b_h  I  S_v - m_v  S_v $$ 
$$\dot I_v = b_h I S_v - m_v  I_v $$ 
    




## What to do {#shinytab3}

### Notes

**The tasks below are described in a way that assumes everything is in units of MONTHS (rate parameters, therefore, have units of inverse months). If any quantity is not given in those units, you need to convert it first (e.g. if it says a year, you need to convert it to 12 months).**

_Some of the simulations might take a few seconds to run. Be patient._

```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 9 #number of tasks
nrecord = 13 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Set the model parameters such that it corresponds to the following setting: 1000 susceptible hosts and vectors, 1 initially symptomatic host, no infected vectors or pathogen in the environment.Simulation duration approximately 5 years. Assume that only symptomatic individuals transmit, at a rate of 0.002. All other transmission rates should be 0. Assume that the duration of the symptomatic period is 1 month long, the duration of the presymptomatic period is half a month long. Assume that there are no asymptomatic infections. You can, therefore, set the rate of recovery of asymptomatics to anything, it doesn't matter because nobody will be asymptomatic. Assume that no environmental shedding and decay occurs. Assume nobody dies due to disease, and immunity does not wane. Assume that there are no births or natural deaths occurring. With parameters set to correspond to the scenario just described, run the simulation and ensure you get a single outbreak with ~20% susceptibles left at the end."

# Record for task 1
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Let's now assume that 50% of infected hosts are asymptomatic (_f_) and that the duration of the asymptomatic stage (_gA_) is the same as the symptomatic stage (_gI_). Also, assume that asymptomatic infected are half as infectious as symptomatic infected, and that pre-symptomatic are as infectious as symptomatic. Run the simulation, you should get an outbreak with around ~11% susceptibles left.
B) Now, we will envision a scenario where we can only detect and quarantine individuals that show symptoms. Assume that quarantining reduces infectiousness by half. Implement such a scenario, run the simulation.
C) Now, we will envision a scenario where we can quarantine everyone who has become infected, independent of symptom status. Implement such a scenario, again run the simulation and record the number of susceptibles left at the end."

# Record for task 2
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("The number of susceptibles left at the end (Part B)",
                    "The number of susceptibles left at the end (Part C)")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report rounded integer",
                    "Report rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  


# Task 3
tid = 3
alltasks[tid,"TaskText"] = "A) Change settings back as they were in task 2 part A. Now assume that we can administer a drug. It will likely only be given to symptomatic individuals. First assume that the drug reduces the infectiousness of symptomatic individuals by half. Run the simulation. 
B) Now assume that the drug _also_ reduces the duration of the symptomatic period from a month (30 days) to 20 days. Run the simulation."

# Record for task 3
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("The number of susceptibles left at the end (Part A)",
                    "The number of susceptibles left at the end (Part B)")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report rounded integer",
                    "Report rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Of course, intervention strategies are best if they reach all that transmit. Let's assume now that we still have a drug that targets symptomatic individuals, but that asymptomatic and presymptomatic individuals don't transmit.Set the model such that only symptomatic individuals transmit, at rate 0.004 and duration of the symptomatic period to 1 month. Nobody else transmits. Everything else should be as in task 2. You should get an outbreak of the same size as in task 1.
B) Now assume that the drug reduces the duration of the symptomatic period from a month (30 days) to 20 days. Run the simulation. 
C) Now assume that the drug also cuts infectiousness by half. Run the simulation."

# Record for task 4
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("The number of susceptibles left at the end (Part B)",
                    "The number of susceptibles left at the end (Part C)")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report rounded integer",
                    "Report rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) Set everything as in task 1. Then turn on environmental shedding by symptomatic individuals and decay both at rates of 1 per month. Run the simulation."

# Record for task 5
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("The number of susceptibles left at the end")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 6
tid = 6
alltasks[tid,"TaskText"] = "A) Set the rate of transmission from the environment to susceptible hosts to 0.002. Run the simulation.
Now, turn off the rate of direct transmission between hosts. Make sure you still get an outbreak. 
B) Let's assume different interventions that affect the environmental transmission. First, we consider an intervention that leads to a 50% increased pathogen clearance from the environment. Run the simulation.
C) Now instead of faster clearance, assume a 50% reduced rate of infection from the environment. Run the simulation."

# Record for task 6
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("The number of susceptibles left at the end (Part B)",
                    "The number of susceptibles left at the end (Part C)")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recnote = c("Report rounded integer",
                    "Report rounded integer")
reclist$recfuzzy = c(0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task  

# Task 7
tid = 7
alltasks[tid,"TaskText"] = "A) Now we'll switch to vector-borne transmission. Set everything as in task 1, introduce 1 infected vector. Assume that transmission between hosts does not occur. Set transmission from host to vector and vector to host to 0.001. Assume that vectors (say mosquitoes) live for half a month. Set birth rate such that vector population balances at 1000. Run the simulation.
B) Double host-vector and vector-host transmission rates. Now consider some vector control measures. Assume we sprayed against mosquitoes and it reduced the population size by 90%. Set the initial vector population to that value. Run the simulation."

# Record for task 7
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("The total number infected at the end (Part B)")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 8
tid = 8
alltasks[tid,"TaskText"] = "A) Instead of killing vector populations, we now consider a reduction in transmission, e.g. due to the use of bed nets. Set everything as in task 7 part A. Assume that an intervention only reduces transmission _to_ vectors by half. Run the simulation, observe. Now assume that an intervention only reduces transmission _from_ vectors by half. Run the simulation, observe. Finally, assume that the intervention reduces transmission both to and from vectors by half. Run the simulation, observe."

# Record for task 8
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

# Task 9
tid = 9
alltasks[tid,"TaskText"] = "A) Keep exploring. The model has many more parameters that you can change, e.g. allowing births and deaths and waning immunity for hosts, or ID that transmit through multiple routes at the same time (e.g. Zika virus). Investigate how different control strategies work under different scenarios."

# Record for task 9
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```



## Further Information {#shinytab4}


* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* For some additional reading, see e.g. [@klepac13, @klepac15, @kirsch17, @tognotti13].

### References

