---
title: ID Patterns - Practice
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: references.bib
---


```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
This app allows you to explore a model that tracks the same compartments as the _Characteristics of ID_  model. If you haven't yet explored the _Characteristics of ID_ model, I suggest you try out that one first. The model for this app adds a few more processes. It includes natural births and deaths of hosts, seasonal variation in transmission, and waning immunity.

Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.

## The Model {#shinytab2}

### Model Overview

This model has the same compartments as the _Characteristics of ID_ model:  

* **S** - uninfected and susceptible individuals 
* **P** - individuals who are infected and do not yet show symptoms. Those individuals can potentially be infectious
* **A** - individuals who are infected and do not show symptoms. Those individuals can potentially be infectious
* **I** - individuals who are infected and show symptoms. Those individuals are likely infectious, but the model allows to adjust this, including no infectiousness.
* **R** - recovered/removed individuals. Those individuals have recovered and are immune. They can loose their immunity in this model.
* **D** - individuals who have died due to the disease.


We include the following processes in this model:

* Susceptible individuals (S) can become infected by pre-symptomatic (P), asymptomatic (A) or symptomatic (I) hosts. The rates at which infections from the different types of infected individuals (P, A and I) occur are governed by 3 parameters, _b~P~_, _b~A~_, and _b~I~_. 
* All infected individuals first enter the presymptomatic stage. They remain there for some time (determined by rate _g~P~_, the inverse of which is the average time spent in the presymptomatic stage). A fraction _f_ of presymptomatic hosts move into the asymptomatic category, and the rest become symptomatic infected hosts.
* Asymptomatic infected hosts recover after some time (specified by the rate _g~A~_). Similarly, the rate _g~I~_ determines the duration the symptomatic hosts stay in the symptomatic state. For symptomatic hosts, two outcomes are possible. Either recovery or death. The parameter _d_ determines the fraction of hosts that die.
* Recovered individuals are initially immune to reinfection. They can loose their immunity at rate _w_ and return to the susceptible compartment. 
* New susceptibles enter the system/model at a fixed rate _n_. From each compartment (apart from the dead compartment), hosts "leave" after some time (i.e. they die) at rate _m_. The inverse of this rate is the average lifespan of a host. 
* The rates of transmission, _b~P~_, _b~A~_ and _b~I~_ can vary seasonally/annually, modeled as a sinusoidal function. The strength of this seasonal variation is controlled by the parameter _s_.

Note that we only track people that die due to the disease in our $D$ compartment. All hosts dying due to other causes just "exit the system" and we don't further keep track of them (though we could add another compartment to "collect" and track all individuals who died from non-disease-related causes.)


### Model Implementation
The flow diagram and equations describe the model implemented in this app:

```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```



$$b_P^s = b_P(1+s \sin(2\pi t / T))$$
$$b_A^s = b_A(1+s \sin(2\pi t /T))$$
$$b_I^s = b_I(1+s \sin(2\pi t /T))$$
$$\dot S = n - S (b_P^s P + b_A^s A + b_I^s I) + wR - m S $$ 
$$\dot P = S (b_P^s P + b_A^s A + b_I^s I) - g_P P - m P$$ 
$$\dot A = f g_P P - g_A A - m A$$ 
$$\dot I = (1-f) g_P P - g_I I - m I $$ 
$$\dot R = g_A A + (1-d) g_I I - wR - m R$$
$$\dot D = d g_I I $$

Since we do not track people dying due to non-disease causes, all the "m - arrows" are not pointing to another compartment, instead of those individuals just "leave the system". Similarly new susceptibles enter the system (are born) from "outside the system".

Also note that the transmission rates, _b~I~_, can be time varying as described above. The parameter _T_ is set depending on the time units chosen for the model. For example if you want to run the model in units of days, the underlying simulation code will set _T=365_, similarly, for weeks it will be _T=52_. This ensures that the seasonal variation always has a period of a year.



## What to do {#shinytab3}

**The tasks below are described in a way that assumes everything is in units of __MONTHS__ (rate parameters, therefore, have units of inverse months). If any quantity is not given in those units, you need to convert it first (e.g. if it says a year, you need to convert it to 12 months).**

_Some of the simulations might take a few seconds to run. Be patient._

```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 11 #number of tasks
nrecord = 19 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1

# this is a list where all the items that need to be recorded are stored
reclist = list()



# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Run the simulation with A population size, _S~0~_, of 1000, , 1 initially infected and presymptomatic host, simulation duration (approximately) 10 years. Assume that only symptomatic individuals transmit, at rate _b~I~_ = 0.002 and that there is no seasonal variation. Assume that the duration of the presymptomatic, asymptomatic and symptomatic periods are all 5 days long. (Hint: The parameters _g~P~_, _g~A~_, and _g~I~_ are the inverses of these periods.) Assume that the duration of the symptomatic period is 1 month long. (Hint: The parameter _g~I~_ is the inverse of this period.) Assume that the duration of the presymptomatic period is approximately 6 days long. (Make sure you convert units correctly. You can assume that a month is roughly 30 days.) Assume that there are no asymptomatic infections. You can, therefore, set the rate of recovery of asymptomatics, _g~A~_ to anything, it doesn't matter because nobody will be asymptomatic. Assume nobody dies due to disease, immunity does not wane, no births, and no natural deaths. If you did it correctly, you should get a single outbreak with around 203 susceptibles left at the end."

# Record for task 1
nrec = 1 # number of items to record
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Next, turn on births and deaths. It's always good to check parts of a model. Here, we want to first look at births and deaths in the absence of disease. Therefore, set initial number of presymptomatic to 0, keep the number of susceptible at 1000. Set the natural mortality rate, _m_, to a value that corresponds to an average lifespan of 41 years. Recall that the unit of _m_ needs to be in 1/months. You need to convert lifespan to months, then take the inverse to get the death rate. Round to the first significant digit (i.e. 0.00X). Set the birth rate to _n=4_ per month. Also increase the simulation duration to 1000 years. (Depending on the speed of your computer, it will take several seconds for the simulation to finish).
B) It is possible to compute the steady state population size of susceptibles in the absence of infection, by setting the differential equation for S to 0 and solving for S as a function of birth and death parameters. Compute this equation (you should get something like _S=.._ with the right side involving parameters _m_ and _n_. Then stick in the values for the parameters from the previous task and compute _S_ at steady state. Make sure it's the same you get from the simulation (once it has settled down to a steady state)."

# Record for task 2
nrec = 1 # number of items to record
reclist$rectext = c("Total number susceptibles at steady state")
reclist$recnote = c("Report the rounded integer")
reclist$rectype = c("Numeric")
reclist$recfuzzy = c(1)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


# Task 3
tid = 3
alltasks[tid,"TaskText"] = "A) Now set _S~0~_ = 3000, keep everything else unchanged. Run the simulation and note what you get for the final, steady-state population size. Try out different values for the initial number of susceptibles and see what happens. 
B) Set population size back to 1000, now try out different values for the birth and death rates, specifically, set _n=2_ and _m=0.002_ and _n=6_ and _m=0.003_. For each setting, think about what you expect to see, based on your intuition and based on the theory/math, i.e. the equation from the previous task, and compare with the simulation results. "

# Record for task 3
nrec = 3 # number of items to record
reclist$rectext = c("Total number susceptible at steady state (part A; _S~0~=3000_)",
                    "Total number susceptible at steady state (part B; _n=2_ and _m=0.002_)",
                    "Total number susceptible at steady state (part B; _n=6_ and _m=0.003_)")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric")
reclist$recfuzzy = c(0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


# Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Now, set the birth and death rate as in task 2 and also adjust the population size to the steady state value you found in task 2. We do that because we don't want a mixing of underlying population growth/decline dynamics on top if the ID dynamics. That would make it more complicated to understand what's going on. (You get to do that later). Ensure that if you run the simulation now, nothing changes. Now introduce an infected individual (_P0=1_), with other values as set in task 1. If needed, adjust the simulation duration to see the initial pattern and ensure everything reaches steady state. What do you expect to see? Run the simulation, compare expectations with results."

nrec = 1 # number of items to record
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) It is possible to compute the steady state values for the model variables at the endemic scenario, i.e. when the number of infected is non-zero. This computation is a bit messy for the model above because it has the extra _P_ and _A_ states that make the math more complicated (you can try for yourself as a challenge after you've done the simple model we'll do next.) To keep the math simpler, we consider a simplified approximation of the above model. Since the way we have set our parameters so far, individuals spend a very short time in _P_ compared to _I_, so we can ignore the _P_ compartment and assume that individuals directly go from _S_ to _I_. Further, for our parameter choices, nobody enters _A_ and _D_, so we can ignore those compartments too. This produces the following simpler SIR model:

$$ dS/dt = n -   b_I IS - m S $$ 
$$ dI/dt = b_I SI - g_I I - m I $$ 
$$ dR/dt =  g_I I  - m R $$

Following the procedure above for computing the steady state for _S_ in the absence of a pathogen, now compute the endemic steady state by setting the left side of the equations to 0 and solving for the variables _S_, _I_ and _R_ as functions of model parameters only. You should get _S=(g~I~ + m)/b~I~ _ and similar equations for _I_ and _R_. For more on steady state calculations, see e.g. [@vynnycky10] or [@keeling08] (but note that each of those references uses their own notation which is not the same as used here.)
B) Let's compare the model simulations with the equations you found. Substitute the parameter values from the task above into the equations and confirm that you get the approximately the same results as you get from running the model to steady state."

# Record for task 5
nrec = 1 # number of items to record
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 6
tid = 6
alltasks[tid,"TaskText"] = "A) The patterns we've seen so far are oscillations/cycles (i.e. repeated outbreaks) that eventually settle down to a steady state. The oscillations are what is often referred to as _intrinsic_, purely produced by the interplay between depletion of susceptibles due to infection and replenishment due to birth. Waning immunity is another mechanism of resource replenishment. To investigate that, set everything back as in task #1 (i.e. no more births and deaths). Do a quick run to ensure you get a single outbreak.
B) Now, assume that immunity wanes within on average 10 months and set simulation time for 100 years. Think about your expectations, run the simulation and check with what you expected. Double the _duration_ of immunity. What do you expect? What do you see?"

# Record for task 6
nrec = 2 # number of items to record
reclist$rectext = c("Total number infected at the end (part B; 10 Months)",
                    "Total number infected at the end (part B; 20 Months)")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$rectype = c("Numeric",
                    "Numeric")
reclist$recfuzzy = c(0,
                     0)

allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 7
tid = 7
alltasks[tid,"TaskText"] = "A) Take the SIR model from above, add the terms describing waning immunity, and use this new set of equations to again solve for the steady states of the model. Then compare the steady state values from the equations with those from the simulations. The equations will directly let you see which variables do and don't change if waning immunity changes. This allows you to confirm what you found in the previous task. This task is challenging! The equations are getting big and ugly. You might want to consider using software that can solve such equations analytically. The free [Maxima software](http://maxima.sourceforge.net/) might be a good option."

# Record for task 7
nrec = 1 # number of items to record
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)

allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 8
tid = 8 
alltasks[tid,"TaskText"] = "A) Now combine births and deaths with waning immunity. Explore how the two different mechanisms interact and influence the observed ID dynamics. Set everything as in task 1. Then, set waning immunity to 10 months, birth rate _n=10_, natural death rate _m=0.005_, and simulation time of 600 months."

# Record for task 8
nrec = 1 # number of items to record
reclist$rectext = c("Total number of recovered at end of outbreak")
reclist$rectype = "Numeric"
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)

allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 9
tid = 9
alltasks[tid,"TaskText"] = "A) So far, the cycles were due to the intrinsic dynamics of the system. Now we'll explore what happens if some outside phenomenon - e.g. the weather - influences some model parameters. Generally, when running the kind of models we investigate here, the variables change (and each gets its own equation) and the parameters remain fixed during the simulation. Here, we are looking at a model where this is not the case. The model allows you to let the transmission rate parameters vary sinusoidally with a period of 1 year. (It would also be possible to instead give each changing parameter their own equation, but that would be more complicated here). Set everything as in task #1, do a quick run to make sure you only get one outbreak. Now run the model with increasing seasonality by setting s = 0.001, 0.01, 0.1, and 1. How many outbreaks do you observe? What else new is going on? Think about the system and formulate your expectations, then run the model to check."

# Record for task 9
nrec = 4 # number of items to record
reclist$rectext = c("Total number of outbreaks (s = 0.001)",
                    "Total number of outbreaks (s = 0.01)",
                    "Total number of outbreaks (s = 0.1)",
                    "Total number of outbreaks (s = 1)")
reclist$recnote = c("Report integer",
                    "Report integer",
                    "Report integer",
                    "Report integer")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric",
                    "Numeric")
reclist$recfuzzy = c(0,
                     0,
                     0,
                     0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec



# Task 10
tid = 10
alltasks[tid,"TaskText"] = "A) Assume now that immunity wanes within 10 months. Also, set the strength of seasonal variation to 0.5. Everything else as before. Think about your expectations. Run the model and compare. Adjust the simulation time to _zoom in/out_ by running for both 120 and 1200 months. Explore how prolonging immunity, i.e. decreasing _w_ to 0.05 and 0.01, changes the patterns you see. Repeat the simulations for the 3 values of the immunity parameter at a lower seasonal variation value of s=0.1. Contemplate the main difference between the pattern you now observe and the ones you've seen so far with no external driver."

# Record for task 10
nrec = 3 # number of items to record
reclist$rectext = c("Total number of outbreaks (_w=0.1_, _s=0.1_, _tmax=1200_)",
                    "Total number of outbreaks (_w=0.05_, _s=0.1_, _tmax=1200_)",
                    "Total number of outbreaks (_w=0.01_, _s=0.1_, _tmax=1200_)")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer",
                    "Report the rounded integer")
reclist$rectype = c("Numeric",
                    "Numeric",
                    "Numeric")
reclist$recfuzzy = c(1,
                     1,
                     1)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


# Task 11
tid = 11 
alltasks[tid,"TaskText"] = "A) Keep exploring! You can let people enter the A and D compartments, have an underlying growing or declining population in the absence of disease and let the disease run on top of it, allow for birth/death, waning immunity and seasonality, etc. There is lots to explore. While it can be fun running simulations with random parameters, I suggest a more deliberate exploration. Before you run a simulation, think about what you expect. Then run and compare results with expectations. If they don't agree, try to figure out what's going on. This way, you will gain more and more insight into how the different components and processes influence the observed outcome - the basics of a systems science approach."

# Record for task 11
nrec = 1 # number of items to record
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)

allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```



## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* For some papers discussing ID patterns and seasonality, see e.g. [@dowell01; @altizer06].

### References



