---
title: Stochastic SIR Model 
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsaide_references.bib
---

```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```



## Overview {#shinytab1}
This app lets you explore a _stochastic_ SIR model. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.

### Learning Objectives
* Get an introduction to stochastic compartmental models
* Understand differences and similarities between stochastic models and their deterministic equivalent
* Understand the concept of reproducibility with random numbers
* Know how the reproductive number affects outbreaks in stochastic models

## The Model {#shinytab2}

### Model Overview

This model tracks susceptibles, infected/infectious and recovered hosts. 
The following compartments are included:  

* **S** - uninfected and susceptible individuals 
* **I** - individuals who are infected and infectious.
* **R** - recovered/removed individuals. Those individuals have recovered and are immune. 


The included processes/mechanisms are the following:

* Susceptible individuals (S) can become infected at rate _b_. 
* Infected hosts recover at rate _g_. 
* New susceptible hosts enter the system (are born) at rate _n_. Natural death occurs at rate _m_.



### Model Implementation
The flow diagram for the model implemented in this app is:



```{r modeldiagram,  fig.cap='',  echo=FALSE}
knitr::include_graphics( system.file(figuredir,appsettings$modelfigname,package=packagename))
```


The deterministic model implemented as set of differential equations is given by the following equations:

$$\dot S = n - bSI - mS$$
$$\dot I = bSI - gI - mI$$
$$\dot R = gI - mR$$

This is almost the same model as the basic SIR model from the introductory app, with the only difference that this model also allows natural births and deaths. In this app, you can run the basic SIR model together with the stochastic one.

The main model we are focused on here is a stochastic implementation of the variables and processes described above. This model is _not_ an ordinary differential equation model. It is instead its stochastic equivalent.
We can specify the model by writing down every possible transition/event/reaction that can occur and their propensities (the propensity multiplied with the
time step gives the probability that a given event/transition occurs). For our model these are the following:

Event type              |  Transitions        |   Propensity      |
----------              |  -----------        |   ----------      |
Infection               |  S => S-1, I => I+1 |   bSI   |
Recovery                |  I => I-1, R => R+1 |   gI           |
Births                  |  S => S+1           |   n          |
Death of susceptible    |  S => S-1           |   mS          | 
Death of infected       |  I => I-1           |   mI          |
Death of recovered      |  R => R-1           |   mR          |



### A note on randomness in computer simulations
This simulation (as well as some of the others) involves using random numbers to introduce stochasticity/uncertainty/noise into the model. This leads to a model that usually more closely reflects the underlying real system. However, in science, we want to be as reproducible as possible. Fortunately, random numbers on a computer are not completely random, but can be reproduced. In practice, this is done by specifying a random number seed, in essence a starting position for the algorithm to produce pseudo-random numbers. As long as the seed is the same, the code should produce the same pseudo-random numbers each time, thus ensuring reproducibility.


## What to do {#shinytab3}

**The tasks below are described in a way that assumes everything is in units of DAYS (rate parameters, therefore, have units of inverse days). If any quantity is not given in those units, you need to convert it first (e.g. if it says a week, you need to convert it to 7 days).**

_Some of the simulations might take a few seconds to run. Be patient._


```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 5 #number of tasks
nrecord = 6 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1

# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Start with 1000 susceptibles, 1 infected, simulation time of 100 days. Transmission rate of 0.0005, duration of infectious period 5 days, no births and deaths, random seed of 100. Run the deterministic simulation multiple times. You should always see the same result.
B) Now run a single stochastic simulation multiple times. What do you expect? What do you see? Change the random seed to 101, then 102. Test how this changes the simulation."

# Record for task 1
nrec = 2 # number of items to record
reclist = list()
reclist$rectext = c("Total number of recovered at end of outbreak with random seed = 101",
                    "Total number of recovered at end of outbreak with random seed = 102")
reclist$rectype = c("Numeric", 
                    "Numeric")
reclist$recnote = c("Report the rounded integer",
                    "Report the rounded integer")
reclist$recfuzzy = c(0,0)

allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec #increment record counter by number of outcomes to record for this task 


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Keep everything as in task 1, with random number seed at 100. Run both simulations and set the number of stochastic simulations to 5. You'll get a plot showing the deterministic model dynamics and 5 runs for the stochastic model (each of the stochastic runs gets consecutive random number seeds, i.e. the first one is 100, the 2nd is 101,...). Try to understand and explain why you see some stochastic simulations lead to an outbreak, while some do not."

# Record for task 2
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("How many simulations produced an outbreak?")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 3 
tid = 3
alltasks[tid,"TaskText"] = "A) Try to understand and explain what the fact that you outbreaks in some of the stochastic simulations but not in others (for exactly the same model parameter values) says about the role of _R~0~_ in stochastic models and how that differs in deterministic models."

# Record for task 3
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) You can mathematically derive an equation linking the probability that an outbreak occurs, _p_, to the reproductive number (see e.g. [@keeling08]). We won't try to do the math here but instead see if we can use the simulation to figure it out. Start with the settings in task 1. Run the stochastic model for b = 0.0002, 0.0003, 0.0004, 0.0005, 0.0006, 0.001, 0.002. For each value, compute R~0~, run the model 100 times, with max time 300 and record the number of times you get an outbreak. Plot the relationship (either by hand or using the computer) between the probability of getting an outbreak and _R~0~_."

# Record for task 4
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Is there a postive or negative relationship between R~0~ and the probability of getting an outbreak?")
reclist$recnote = c("Report 'positive' or 'negative'")
reclist$rectype = c("Character")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) There is a similar relation between outbreak probability an the initial number of infected individuals. Let's see if we can use the model to determine it. Follow the steps of the previous task. Keep _b_ fixed at 0.0003, vary the initial number of infected individuals, _I~0~_, from 1 to 10. Record the fraction of outbreaks you got for each _I~0~_. Plot the relationship (either by hand or using the computer) between the probability of getting an outbreak and _I~0~_."

# Record for task 5
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Is there a positive or negative relationship between the initial number of infected and the probability of getting an outbreak?")
reclist$recnote = c("Report 'positive' or 'negative'")
reclist$rectype = c("Character")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```



## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* For some further reading concerning stochastic dynamics of ID, see e.g. [@lloyd-smith05a; @black66; @keeling97].

### References

