---
title: Vector Borne Transmission 
output:
  html_document:
    theme: null
    highlight: null
    fig_retina: null
    fig_caption: true
    mathjax: default 
    keep_md: false
bibliography: dsaide_references.bib
---

```{r, include = FALSE}
#*************************************
#general setup to define package and get path locations
#all paths are inside the package and retrieved with system.file
packagename = "DSAIDE"
helperdir = "helperfunctions"
mbmodeldir = "mbmodels"
figuredir = "media"
appdocdir = "appinformation" 
#*************************************
#Note: for this to process/knit, several helper functions need to be available (sourced) first
#those are in the inst/helperfunctions folder
#Note: in general, the "processing-script.R" in the docsfordevelopers should be used to produce the html docs
#manual knitting of each doc only during development/testing
#*************************************
files_to_source = list.files(system.file(helperdir,package = packagename),full.names=TRUE)
sapply(files_to_source, source) #sourcing needs to happen inside each Rmd file since knitr starts a new environment
#load the settings file for the current app 
#so we can automatically include figure, list the functions in the further information section
#and use other information specific to the current app for the task table generation
currentrmdfile = knitr::current_input() 
appsettings = get_settings(currentrmdfile,appdocdir,packagename)
```


## Overview {#shinytab1}
This app allows you to explore a simple model for vector-borne transmission. Read about the model in the "Model" tab. Then do the tasks described in the "What to do" tab.


## The Model {#shinytab2}

### Model Overview

This is a simple 2 species (hosts and vectors) model, using the basic SIR framework as follows.

The model has the following compartments:  

* **Sh** - uninfected and susceptible host 
* **Ih** - infected/infectious hosts
* **Rh** - recovered/removed hosts
* **Sv** - susceptible vectors
* **Iv** - infected/infectious vectors

We assume that once a vector is infected, it stays infected until it dies. Therefore, recovered vectors are not included in the model.


The processes being modeled are:

* Susceptible host can become infected by contact with infected vectors at rate _b~1~_ 
* Susceptible vectors can become infected by contact with infected hosts at rate _b~2~_ 
* Infected hosts recover after some time (specified by the rate _g_). 
* New susceptible vectors are born at a rate _b_. Susceptible and infected vectors die at rate _n_. The inverse of this rate is the average lifespan. 
* Recovered hosts lose their immunity at rate _w_.

We envision hosts to be humans and vectors to be e.g. mosquitoes. We further assume the time scale for the model is short enough that we can ignore births and deaths for human hosts but not vectors. This is likely a decent approximation as long as our simulation time is only a few years.


### Model Implementation
The flow diagram and the set of ordinary differential equations (ODE) which are used to implement this model are as follows:

```{r modeldiagram,  fig.cap='Flow diagram for this model. ',  echo=FALSE}
knitr::include_graphics(here::here('inst/media',appsettings$modelfigname))
```


Model equations:



```{r, echo = FALSE}
if (appsettings$use_mbmodel)
{
  modeldir = system.file("mbmodels", package = packagename) #find path to apps
  mbmodel = readRDS(paste0(modeldir,"/",appsettings$mbmodelname) )
  eqs = DSAIDE::generate_equations(mbmodel)
  eqs
}
```



## What to do {#shinytab3}

**The tasks below are described in a way that assumes everything is in units of months (rate parameters, therefore, have units of inverse months). If any quantity is not given in those units, you need to convert it first (e.g. if it says a year, you need to convert it to 12 months).**


```{r, echo=FALSE, eval=TRUE}
# save all tasks, outcomes, etc. into an R data frame, then print later.
# this data frame is used to automate shiny solutions and quiz generation

# Explanation for each of the columns in the R data frame 

# quizID: MUST BE THE FIRST COLUMN. Used by the grading app. naming structure "dsaide_shorttitle" where shorttitle is from appsettings$shorttitle
# AppTitle: Title used for the app in the dsaide app, from appsettings$apptitle
# AppID: App number in the dsaide menu, from appsettings$appid
# TaskID: Identifies which task the text belongs to 
# TaskText: The text that explains what to do for the task
# RecordID: Identifies if it is the first, second, third,... item to record within the task
# Record: Text explaining what value from the model that nee too be recorded
# Note: Used by students taking quiz. Makes it clear what type of value to enter in "Answers"
# Answers: In master it contains the rigth answer. For students it is where they record the recorded value
# Fuzzy: The numeric value of a margin (within +/- of "right" answer) for questions where the quiz checker should not be too strict 
# Review: Admin use if value is 1 the way the questions is asked may need to be reconsidered. 

ntasks = 8 #number of tasks
nrecord = 8 #number of outcomes to record
#set up task tables
#returned as list for generality, take apart
tablist <- make_tasktable(ntasks,nrecord,appsettings)
alltasks = tablist[[1]]
allrecord = tablist[[2]]

#this is the running counter for the records which starts at 1 
rc=1


# Task 1
tid = 1
alltasks[tid,"TaskText"] = "A) Set the model parameters so that there are 1000 initially susceptible for both hosts and vectors, 1 initially infected host, no infected vector, simulation duration approximately 10 years. Set the transmission rates, _b~2~_=0.003, and _b~1~_ = 0. Assume that the duration of the infectious period in hosts is half a month (about 2 weeks) long. Assume that vectors live for half a month (about 2 weeks). Set the birth rate such that in the absence of the ID, you get a steady population of 1000 vectors. You should have learned the relation between births, deaths and steady state susceptibles in the _Patterns of ID_ app. If you don't remember, you can figure it out by trying different values for the birth rate (with no infected hosts or vectors present). Turn off waning immunity for now. Run the simulation, observe what you get. You should not see any outbreak happening. Make sure you know why."

# Record for task 1
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


# Task 2
tid = 2
alltasks[tid,"TaskText"] = "A) Now set _b~1~_ and _b~2~_ = 0.003, and rerun the simulation. You should see an outbreak with around 183 susceptible hosts left."

# Record for task 2
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Max number of infected hosts")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 3 
tid = 3
alltasks[tid,"TaskText"] = "A) Play around with _b~1~_ and _b~2~_ and see how different values affect the ID dynamics."

# Record for task 3
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec

### Task 4
tid = 4
alltasks[tid,"TaskText"] = "A) Now set everything back as in task #2 and start with no infected hosts and 1 infected vector. What do you expect to see? Run the simulation to check your expectation.
B) Play around with a different initial number of infected hosts and vectors and different transmission rates. Anything surprising happening? Do you understand why you see what you see?"

# Record for task 4
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Max number of infected hosts (Part A)")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 5
tid = 5
alltasks[tid,"TaskText"] = "A) Turn off any infection process by setting the number of initial infected hosts and vectors to 0. Adjust the birth rate such that the vector population settles at 5000. Set the maximum simulation time to 12 months. Note the time it takes for the vector population to go from the initial 1000 to 5000. This rapid increase can be a problem for vector control measures."

# Record for task 5
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("The number of months it takes for Sv to be 4000")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 6
tid = 6
alltasks[tid,"TaskText"] = "A) Turn infection process back on by having 1 infected host. Set simulation time back to 10 years. Everything else as above. Run the simulation, contemplate the dynamics you see. You have births and deaths now, and usually with those, you can get multiple outbreaks and an endemic state. Why not here?"

# Record for task 6
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 7
tid = 7
alltasks[tid,"TaskText"] = "A) Assume host immunity wanes after an average of 2 months. What do you expect to see? Run simulations to check your expectations. You should get a steady endemic state with a constant, non-zero number of infected hosts and vectors. The number of infected vectors at steady, endemic state should be around 1052."

# Record for task 7
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Final number of suscepitble hosts at steady state")
reclist$rectype = c("Numeric")
reclist$recnote = c("Report the rounded integer")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec


### Task 8
tid = 8
alltasks[tid,"TaskText"] = "A) Keep exploring. Try to figure out how the different parameters influence the dynamics, specifically the (lack of) sustained cycles."

# Record for task 8
nrec = 1 # number of items to record
reclist = list()
reclist$rectext = c("Nothing")
reclist$rectype = "None"
reclist$recnote = c("")
reclist$recfuzzy = c(0)
allrecord <- fill_tasktable(allrecord,tid,rc,nrec,reclist) #the allrecord table is updated with the new entries
rc = rc + nrec

```


```{r echo=FALSE}
#save the fully filled task table to a tsv file
save_tasktable(alltasks,allrecord,appsettings)
```


```{r, echo=FALSE, results='asis'}
# Take all the text stored in the table and print the tasks and items to record
write_tasktext(alltasks,allrecord)
```


## Further Information {#shinytab4}

* This app (and all others) are structured such that the Shiny part (the graphical interface you see and the server-side function that goes with it) calls an underlying R script (or several) which runs the simulation for the model of interest and returns the results.
* For this app, the underlying function(s) running the simulation are called ``r appsettings$simfunction``. You can call them directly, without going through the shiny app. Use the `help()` command for more information on how to use the functions directly. If you go that route, you need to use the results returned from this function and produce useful output (such as a plot) yourself. 
* You can also download all simulator functions and modify them for your own purposes.  Of course to modify these functions, you'll need to do some coding.
* For examples on using the simulators directly and how to modify them, read the package vignette by typing `vignette('DSAIDE')` into the R console.
* Some more information on vector-borne diseases and modeling can be found in [@kilpatrick12].


### References

